---
title: "W271 Group Lab"
subtitle: "Bike share demand"
author: "Please fill in with your names."
output: bookdown::pdf_document2
fontsize: 11pt
geometry: margin=1in
---

# Introduction (5 points)

**Complete the following task. In your final submission, please remove this question prompt so that your report reads as a report.**

The `SeoulBikeData` data set measures the number of bike rentals (bikers) per hour in Seoul, South Korea, from December 1, 2017, to November 30, 2018. There are 8760 records/observations and 14 variables in this data set. The ‘Seoul Bike Sharing Demand Data Set’ can be downloaded from UC Irvine Machine Learning Repository: <https://archive.ics.uci.edu/ml/datasets/Seoul+Bike+Sharing+Demand>

Initiate your analysis by formulating a hypothesis related to bike share demand that can be tested using a regression model. This hypothesis aims to provide decision-makers with insights into the factors influencing bike usage and demand, facilitating the optimal availability of rental bikes to the public when needed.

Hypothesis: Seasonal, Daily (holiday, weekday, weather) and Hour

# Data (20 points)

## Description (5 points)

```{r}
library(tidyverse)
library(knitr)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(lubridate)
library(car)
```

**Complete the following task. In your final submission, please remove this question prompt so that your report reads as a report.**

Describe the data that you are using. How is this data generated, what is the sampling process that brought it to your availability, etc. If it is helpful, you might describe the population (i.e. the Random Variables) that exist and how samples are produced from these random variables.

```{r}
library(readr)
bike_demand <- read_csv("../data/SeoulBikeData.csv", locale = locale(encoding="latin1"))
# rename
names(bike_demand) <- c("date", "rented_bike_count", "hour", "temperature_c", "humidity",
               "wind_speed_m_s", "visibility_10m", "dew_point_temp_c", 
               "solar_radiation_mj_m2", "rainfall_mm", "snowfall_cm", 
               "season", "holiday", "functioning_day")
head(bike_demand)
```

Sampling process

## EDA (15 points)

**Complete the following task. In your final submission, please remove this question prompt so that your report reads as a report.**

Examine the dataset and investigate anomalies, missing values, and convert variables when necessary for future analysis.

```{r}
print(summary(bike_demand))
str(bike_demand)

# Check nans - zero
colSums(is.na(bike_demand))

# Check duplicates - zero
bike_demand %>%
  group_by(date, hour) %>%
  summarise(count = n(), .groups = "drop") %>% 
  arrange(desc(count))

# Check if anormalies in bike demand - right skewed
bike_demand %>% 
  filter((functioning_day == "Yes") & (rented_bike_count==0))

# Data transformation, extract `day`, `month` and `year` from `date` column
bike_demand$date <- as.Date(bike_demand$date, format = "%d/%m/%Y")
bike_demand <- bike_demand %>%
  mutate(
    day = day(date),
    month = month(date),
    year = year(date),
    is_weekday = ifelse(wday(date) %in% c(1, 7), "no", "yes")
  )

# Add a feature for rush hour
bike_demand <- bike_demand %>%
  mutate(is_rush_hour = ifelse(hour %in% c(6:9, 17:19), "yes", "no"),
         is_day = ifelse(hour %in% c(5:19), "yes", "no"))

# Add log transformation for `rented_bike_count`, `rainfall_mm` and `snowfall_cm`
bike_demand <- bike_demand %>%
  mutate(log_bike_count = log(rented_bike_count),
         log_rainfall = log1p(rainfall_mm),
         log_snowfall = log1p(snowfall_cm))

# Filter out non-functional days
bike_demand_filtered <- bike_demand %>% 
  filter(functioning_day == "Yes")
```

Conduct a thorough EDA of the data set, giving special attention to the relationships relevant to your hypothesis.

-   This should include both graphical and tabular analysis as taught in this course.
-   Since the report has a page-limit, you will have to be selective when choosing visuals to illustrate your key points, associated with a concise explanation of the visuals.

### Bike Demand Visualization

-   log transformation can be used to normalize the fat tail in `rented bike count`

```{r}
# hist(bike_demand$rented_bike_count)
plot_bike_cnt1 <- bike_demand_filtered %>% 
  ggplot(aes(x = rented_bike_count)) +
  geom_histogram(aes(y = after_stat(count / sum(count))), width = 2, fill = "#999999", color = "white") +
  ggtitle("Distribution of rented bike count") + 
  xlab("rented bike count") +
  ylab("percentage") +
  theme_minimal()

plot_bike_cnt1

plot_bike_cnt2 <- bike_demand_filtered %>% 
  ggplot(aes(x = rented_bike_count)) +
  geom_boxplot() +
  ggtitle("Distribution of rented bike count") + 
  xlab("rented bike count") +
  ylab("")

grid.arrange(plot_bike_cnt1, plot_bike_cnt2, nrow = 1, ncol = 2)
```

### Hour of the Day - weekday + rush hour (how to extract rush hour?)

-   On average, hourly bike demand is similar on weekdays vs weekends

-   Except for rush hours on weekdays, which see significant increase in bike demand

```{r}
p1 <- bike_demand_filtered %>%
  group_by(hour, is_weekday) %>%
  summarise(avg_rented = mean(rented_bike_count, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = hour, y = avg_rented, color = factor(is_weekday))) +
  geom_line(size = 1.2) +
  scale_color_manual(
    values = c("yes" = "#6688BB", "no" = "#ffc000"),
    labels = c("yes"="Weekday", "no" = "Weekends"),
    name = "Day Type"
  ) +
  labs(
    title = "Average Rented Bikes by Hour",
    x = "Hour of the Day",
    y = "Average Rented Bikes"
  ) +
  theme_minimal()

p1

p2 <- bike_demand_filtered %>%
  group_by(is_rush_hour, is_weekday) %>%
  summarise(avg_rented = mean(rented_bike_count, na.rm = TRUE)) %>%
  ggplot(aes(x = factor(is_rush_hour), y = avg_rented, fill = factor(is_weekday))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
  values = c("yes" = "#6688BB", "no" = "#ffc000"),
  labels = c("yes"="Weekday", "no" = "Weekends"),
  name = "Day Type") +
  labs(
    title = "Rush Hours vs Non-Rush Demand",
    x = "Rush Hour",
    y = "Average Rented Bikes"
  ) +
  theme_minimal()
p2
grid.arrange(p1, p2, nrow = 1, ncol = 2)
```

### On Daily Basis

#### Day of the month - doesn't seem to impact bike demand

```{r}
bike_demand_filtered %>%
ggplot(aes(x = day, y = rented_bike_count, color=season)) +
geom_point(alpha = 0.5)
```

#### Holiday - less bike demand on holidays, except for autumn

```{r}
p3 <- bike_demand_filtered %>%
  group_by(season, holiday) %>%
  summarise(avg_rented = mean(rented_bike_count, na.rm = TRUE)) %>%
  ggplot(aes(x = season, y = avg_rented, fill=factor(holiday))) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
  values = c("Holiday" = "#ffc000", "No Holiday" = "#6688BB"),
  labels = c("Holiday"="holiday", "No Holiday" = "non-holidays"),
  name = "Day Type") +
  labs(
    title = "Average Rented Bikes by Season",
    x = "Season",
    y = "Average Daily Rented Bikes"
  )
p3

```

#### Weather

1.  Temperature & humidity
    -   temperature is highly correlated with season, both have a big impact on bike demand

    -   humidity has a quadratic relationship with bike demand

```{r}
p4 <- bike_demand_filtered %>% 
  ggplot(aes(x = temperature_c, y = rented_bike_count, color = factor(season))) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Temperature vs Bike Rentals by Season",
    x = "Temperature (°C)",
    y = "Rented Bike Count"
  ) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  theme_minimal()

p5 <- bike_demand_filtered %>% 
  ggplot(aes(x = humidity, y = rented_bike_count, color = factor(season))) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Humidity vs Bike Rentals by Season",
    x = "Humidity",
    y = "Rented Bike Count"
  ) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  theme_minimal()
grid.arrange(p4, p5, nrow = 1, ncol = 2)
```

2.  rainfall & snowfall - both are heavily skewed and negatively impact bike demand

```{r}
p6 <- bike_demand_filtered %>% 
  ggplot(aes(x = log1p(rainfall_mm), y = rented_bike_count, color = factor(season))) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Rainfall vs Bike Rentals",
    x = "Rainfall",
    y = "Rented Bike Count"
  ) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  theme_minimal()

p7 <- bike_demand_filtered %>% 
  ggplot(aes(x = log1p(snowfall_cm), y = rented_bike_count, color = factor(season))) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Snowfall vs Bike Rentals",
    x = "Snowfall",
    y = "Rented Bike Count"
  ) +
  geom_smooth(method = "loess", se = FALSE, color = "blue") +
  theme_minimal()
grid.arrange(p6, p7, nrow = 1, ncol = 2)
```

### By Season -

-   `season` has big impact on bike demand(check the `holiday` graph)

-   Compared to `hour`, `weekday`, the impact from `season` is less obvious

```{r}
p8 <- bike_demand_filtered %>%
  filter(is_weekday=="yes") %>%
  group_by(hour, season) %>%
  summarise(avg_rented = mean(rented_bike_count, na.rm = TRUE)) %>%
  ggplot(aes(x = hour, y = avg_rented, color = factor(season))) +
  geom_line() +
  labs(
    title = "Bike Demand by Hour(weekdays)",
    x = "Hour of the Day",
    y = "Average Rented Bikes(hourly)"
  )

p9 <- bike_demand_filtered %>%
  filter(is_weekday=="no") %>%
  group_by(hour, season) %>%
  summarise(avg_rented = mean(rented_bike_count, na.rm = TRUE)) %>%
  ggplot(aes(x = hour, y = avg_rented, color = factor(season))) +
  geom_line() +
  labs(
    title = "Bike Demand by Hour(weekends)",
    x = "Hour of the Day",
    y = "Average Rented Bikes(hourly)"
  )

grid.arrange(p8, p9, nrow = 1, ncol = 2)
```

# Model Development (40 points)

## Poisson regression (10 points)

**Complete the following task. In your final submission, please remove this question prompt so that your report reads as a report.**

Estimate a simple poisson regression model (`model_poisson_1`) that includes variables as linear terms in the model based on what is essential for your hypothesis. Evaluate, using likelihood ratio tests and the statistical significance of each explanatory variable in the model. Also evaluate, using the context and data understanding that you have created in the Data section of this report, the practical significance of each explanatory variable in the model.

```{r}
bike_demand_filtered <- bike_demand_filtered %>%
  mutate(
    season = relevel(factor(season), ref = "Winter"),
    is_rush_hour = relevel(factor(is_rush_hour), ref = "no"),
    is_weekday = relevel(factor(is_weekday), ref = "no")
  )
```

```{r}
model_poisson_1 <- glm(rented_bike_count ~ is_weekday + hour + temperature_c + log1p(rainfall_mm) + log1p(snowfall_cm) + season, data = bike_demand_filtered, family = poisson())
summary(model_poisson_1)
# Likelihood ratio test
library(car)
Anova(mod = model_poisson_1, type = 3, test="LR") #This tests each variable as if it were added last (type=3)
```

Conclusion:

-   rush hour, seasonality(winter particularly) and rainfall make a big impact bike demand

## Model Comparison (10 points)

Expand your initial poisson model through a two-step process:

-   Incorporate additional variables from the dataset to avoid potential omitted variable bias. (`model_poisson_2`)

-   Introduce quadratic terms or interactions when they can potentially explain variations in bike rental demand. For instance, you might consider assessing whether the effect of Hour on bike demand differs between weekdays and weekends. (`model_poisson_3`)

Finally, use three information criteria for the three models (`model_poisson_1`, `model_poisson_2`, `model_poisson_3`) and rank the models based on each criterion. Are the results consistent? Why or why not?

```{r}
model_poisson_2 <- glm(rented_bike_count ~ is_weekday + hour 
                       + is_day + is_rush_hour 
                       + temperature_c + log1p(rainfall_mm) + log1p(snowfall_cm) + season
                       + humidity + wind_speed_m_s + visibility_10m + solar_radiation_mj_m2 + holiday
                       , data = bike_demand_filtered, family = poisson())
anova(model_poisson_1, model_poisson_2, test="Chisq")
summary(model_poisson_2)
```

```{r}
model_poisson_3 <- glm(rented_bike_count ~ is_weekday + hour 
                       + is_day + is_rush_hour 
                       + temperature_c + log1p(rainfall_mm) + log1p(snowfall_cm) + season
                       + humidity + wind_speed_m_s + visibility_10m + solar_radiation_mj_m2 + holiday
                       # + is_rush_hour:is_day # the nans here indicates complete separation between is_rush_hour and is_day
                       + is_day:is_weekday
                       + is_rush_hour:is_weekday
                       + log1p(rainfall_mm) + log1p(snowfall_cm) 
                       + temperature_c
                       + season:temperature_c,
                       data = bike_demand_filtered,
                       family = poisson()
)
summary(model_poisson_3)
```

```{r}
#  poisson fit
library(vcd)
rented_bikes <- bike_demand_filtered$rented_bike_count
rented_bikes_mean <- mean(rented_bikes)
rented_bikes_var <- var(rented_bikes)
rented_bikes_fit <- goodfit(rented_bikes, "poisson")
summary(rented_bikes_fit)
```

```{r}
#  Robust standard errors
library(stargazer)
model_poisson_base_1 <- coeftest(model_poisson_1)
model_poisson_base_2 <- coeftest(model_poisson_2)
model_poisson_base_3 <- coeftest(model_poisson_3)

model_poisson_rse_1 <- coeftest(model_poisson_1, vcov = vcovHC(model_poisson_1, type="HC0"))
model_poisson_rse_2 <- coeftest(model_poisson_2, vcov = vcovHC(model_poisson_2, type="HC0"))
model_poisson_rse_3 <- coeftest(model_poisson_3, vcov = vcovHC(model_poisson_3, type="HC0"))

stargazer(model_poisson_base_1, model_poisson_rse_1, type = "text", title = "Model 1 with Robust SE")
stargazer(model_poisson_base_2, model_poisson_rse_2, type = "text", title = "Model 2 with Robust SE")
stargazer(model_poisson_base_3, model_poisson_rse_3, type = "text", title = "Model 3 with Robust SE")
```
, 
```{r}
#  negative binomial models
library(MASS)
model_nb_1 <- glm.nb(rented_bike_count ~ is_weekday 
                  + hour 
                  + temperature_c 
                  + log1p(rainfall_mm) 
                  + log1p(snowfall_cm) 
                  + season, 
                  data = bike_demand_filtered)

model_nb_2 <- glm.nb(rented_bike_count ~ is_weekday 
                  + hour 
                  + is_day 
                  + is_rush_hour 
                  + temperature_c 
                  + log1p(rainfall_mm) 
                  + log1p(snowfall_cm) 
                  + season
                  + humidity 
                  + wind_speed_m_s 
                  + visibility_10m 
                  + solar_radiation_mj_m2 
                  + holiday, 
                  data = bike_demand_filtered)

model_nb_3 <- glm.nb(rented_bike_count ~ is_weekday
                  + hour 
                  + is_day
                  + is_rush_hour 
                  + temperature_c
                  + log1p(rainfall_mm)
                  + log1p(snowfall_cm)
                  + season
                  + humidity
                  + wind_speed_m_s
                  + visibility_10m
                  + solar_radiation_mj_m2
                  + holiday
                  + is_day:is_weekday
                  + is_rush_hour:is_weekday
                  + log1p(rainfall_mm) 
                  + log1p(snowfall_cm) 
                  + temperature_c
                  + season:temperature_c,
                  data = bike_demand_filtered)

model_nb_1 <- coeftest(model_nb_1)
model_nb_2 <- coeftest(model_nb_2)
model_nb_3 <- coeftest(model_nb_3)

stargazer(model_poisson_base_1, model_nb_1, type = "text", title = "Poisson vs NegBin Model 1")
stargazer(model_poisson_base_2, model_nb_2, type = "text", title = "Poisson vs NegBin Model 2")
stargazer(model_poisson_base_3, model_nb_3, type = "text", title = "Poisson vs NegBin Model 3")


```


**Conclusion:**

-   Weekday vs weekends doesnt make a big difference although it's statistically significant

-   Unless it's weekday rush hour, see rush hour and the interaction term

-   Rainfall reduces the demand, a lot more than snowfall

-   Summer adds demand meaningfully, followed by autumn, surprisingly Spring doesnt add too much

-   Temperature doesnt have a huge influence on average, but different season sees different relationship between temperature and bike demand

## Model Assessment (10 points)

**Complete the following task. In your final submission, please remove this question prompt so that your report reads as a report.**

Use suitable tools like residual or standardized residual plots to assess if your final model is a good approximation of the data. Additionally, check the results for indications of potential overdispersion.

-   How would you address any potential issues, including any overdispersion?

-   Produce an adjusted version of your best model that accounts for issues noted above.

    ```{r}
    residuals<- resid(model_poisson_3)
    fitted_vals <- model_poisson_3$fitted.values

    df_resid <- data.frame(
      fitted_vals = fitted_vals,
      resid = residuals
    )

    # ggplot(df_resid, aes(x = fitted_vals , y = resid)) +
    #   geom_point() +
    #   labs(title = "Residuals vs Fitted values",
    #        x = "Fitted values",
    #        y = "Residuals") +
    #   theme_minimal()

    plot(model_poisson_3, which = 1, main = "Model Fit")
    ```

    **Overdispersion:**

-   Are the days independent?

-   Omitted variables?

-   IID violated?

## Alternative Specification (10 points)

**Complete the following task. In your final submission, please remove this question prompt so that your report reads as a report.**

Using the same set of explanatory variables from your final model, estimate a linear regression model. Examine the summary of the model and interpret the coefficients. Obtain fitted values from both OLS and Poisson regression models and plot them. Discuss which regression model, linear or Poisson, is more suitable for this case and explain why.

# Conclusion (5 points)

**Complete the following task. In your final submission, please remove this question prompt so that your report reads as a report.**

In conclusion, summarize your entire work to leave a lasting impression on the reader. Ensure you:

-   Restate the main hypothesis or research question.

-   Report OLS, Poisson, or any other regressions you estimated.

-   Summarize key findings.

-   Discuss implications.

-   Address limitations.
